---
title: "NYC Shooting Data"
author: "W. Spracklin"
date: "2023-02-02"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

### Importing the Data

This is data from the New York City government that contains shooting data from 2006-2021 in NYC. We will start by getting the url:

```{r get_nyc_shooting_url }
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/"
file <- "rows.csv"
url <- str_c(url_in, file)
```

Now we will download and parse the CSV file from the url created above.

```{r get_nyc_shooting_data }
nyc_shooting_data = read_csv(url)
```

## Tidying, Analyzing, and Visualizing the Data

### Tidying and Summary

There are columns that had to be removed to avoid having noise for the data. Most of the analysis will focus on the victim and the total counts per year of shootings. There are no current columns that capture the total count, so one will be added.

```{r summary}
shooting_data_tidy <- nyc_shooting_data %>%
  select(c('VIC_RACE', 'VIC_AGE_GROUP', 'PERP_SEX', 'VIC_SEX', 'OCCUR_DATE')) %>% 
  mutate(occurrence_date=mdy(OCCUR_DATE)) %>% 
  group_by(occurrence_date) %>% 
  count(occurrence_date)
cleaned_count_data <- shooting_data_tidy %>% mutate(occurrence_date=floor_date(occurrence_date, "year"))
shootings_by_year <- cleaned_count_data %>% group_by(occurrence_date) %>% summarise(number_of_annual_shootings=sum(n))
summary(shootings_by_year)

```

### Analyzing shooting by time of day

In the following graph, we are looking at the time of day when shootings take place. The visual shows that most shootings happen at night and in the early morning, with the weekend having the most occurrences.

```{r nyc_shooting_date_by_time }
shooting_by_time <- nyc_shooting_data %>%
  mutate(OCCUR_DATE=mdy(OCCUR_DATE)) %>%
  mutate(OCCUR_TIME=as.POSIXct(OCCUR_TIME, format="%H:%M")) %>%
  mutate(day=wday(OCCUR_DATE,label=TRUE)) %>%
  group_by(day) %>%
  select(c(day,OCCUR_TIME, OCCUR_DATE))


ggplot(shooting_by_time, aes(x=day, y=OCCUR_TIME)) + geom_point(alpha=1/150, color="red") + xlab("Day") + scale_y_datetime(date_labels='%H') + ylab("Hour of Shooting")

```

### Victims by Sex

In the graph below, a graph is presented that contains a breakdown of the victims' sex and the sex of the perpetrator. The sex was not recorded for some observations, so the data had to be cleaned.

```{r nyc_shooting_data_by_sex }
shooting_by_sex_data <- nyc_shooting_data %>%
group_by(PERP_SEX, VIC_SEX) %>%
select(c('PERP_SEX', 'VIC_SEX')) %>%
  filter( PERP_SEX != "NA" & PERP_SEX != "U" & VIC_SEX != "NA" & VIC_SEX != "U") %>%
summarize(n=n())

ggplot(shooting_by_sex_data, aes(x=PERP_SEX, y=n,fill=VIC_SEX, label=n)) + geom_col() +
     geom_text(size = 2, position = position_stack(vjust = 0.0)) +
     xlab("Perpetrator Sex") + ylab("Number of shootings") +
     scale_fill_discrete(name="Victim Sex")
```

### Victims by Race

In the graph below, the data is showing victims by race, and the race of the perpetrator. Data here was also cleaned because race data was not available for every observation.

Analysis: The data shows that most victims are shot by someone of the same race.

```{r nyc_shooting_data_by_race,echo=TRUE }
shooting_by_race_data <- nyc_shooting_data %>%
  group_by(PERP_RACE, VIC_RACE) %>%
  select(c('PERP_RACE', 'VIC_RACE')) %>%
  filter( PERP_RACE != "NA" & PERP_RACE != "UNKNOWN" & VIC_RACE != "NA" & VIC_RACE != "UNKNOWN") %>%
  summarize(n=n())

ggplot(shooting_by_race_data, aes(x=PERP_RACE, y=n,fill=VIC_RACE)) + geom_col() + xlab("Perpetrator Race") + ylab("Number of shootings")  + scale_fill_discrete(name="Victim Race") + theme_gray(base_size=5)

```
### Victims by Age

In the graph below, data is shown of the victims age and age of the perpetrator. Like above, the data needed to be tidied up to remove both bad and unknown data.

Analysis: From the data, victims are most likely to be shot by someone of their own age group, with the exception of the youngest and the two older groups, with the youngest group most likely to be shot by someone one age older, and the older most likely to be shot by someone from the 25-44 year old age group.  

```{r nyc_shooting_data_by_age, echo=TRUE }


shooting_by_age_data <- nyc_shooting_data %>%
     group_by(PERP_AGE_GROUP, VIC_AGE_GROUP) %>%
     select(c('PERP_RACE', 'OCCUR_DATE', 'PERP_AGE_GROUP','VIC_AGE_GROUP')) %>%
     filter(PERP_AGE_GROUP!='UNKNOWN' & PERP_AGE_GROUP != 'NA' & PERP_AGE_GROUP != '940' & PERP_AGE_GROUP !='1020' & PERP_AGE_GROUP != '224') %>%
     summarize(n=n())

ggplot(shooting_by_age_data, aes(x=PERP_AGE_GROUP, y=n,fill=VIC_AGE_GROUP)) + geom_col() + xlab("Perpetrator Age Group") + ylab("Number of shootings") + scale_fill_discrete(name="Victim Age Group")

```
## Model

In the model below, a linear model is shown for the total shootings from the year 2006 to 2021.

Analysis: Although there was a spike in shootings in 2020 and 2021, the overall trend of shootings is declining. Being that the data only contained shooting data, and no reason for the shootings, it is difficult to say from the data as to what caused the recent spike. It is possible that economic consequences of the COVID19 pandemic might be a factor.

```{r nyc_shooting_data_model, echo=TRUE }
counts <- nyc_shooting_data %>% select(c('OCCUR_DATE')) %>% mutate(odate=mdy(OCCUR_DATE)) %>% group_by(odate) %>% count(odate)
cleaned_count_data <- counts %>% mutate(odate=floor_date(odate, "year"))
shootings_by_year <- cleaned_count_data %>% group_by(odate) %>% summarise(shooting_count=sum(n))
mod_all_shootings <- lm( shooting_count ~ odate, data=shootings_by_year)
summary(mod_all_shootings)
shootings_by_year_pred <- shootings_by_year %>% mutate(pred=predict(mod_all_shootings, shootings_by_year))
ggplot(shootings_by_year_pred, aes(x=odate, y=shooting_count)) + geom_line() + geom_line( color="blue", data=shootings_by_year_pred, aes(x=odate, y=pred)) + xlab("Year") + ylab("Number of Shooting Victims")
```

In this model, the above model is broken down by the race of the victims.
```{r nyc_shooting_data_model_trends, echo=TRUE}
counts <- nyc_shooting_data %>% select(c('OCCUR_DATE')) %>% mutate(odate=mdy(OCCUR_DATE)) %>% group_by(odate) %>% count(odate)
cleaned_count_data <- counts %>% mutate(odate=floor_date(odate, "year"))
shootings_by_year <- cleaned_count_data %>% group_by(odate) %>% summarise(shooting_count=sum(n))

counts_by_race <- nyc_shooting_data %>% select(c('OCCUR_DATE','VIC_RACE')) %>% mutate(odate=mdy(OCCUR_DATE)) %>% group_by(odate, VIC_RACE) %>% count(odate)

cleaned_counts_by_race_data <- counts_by_race %>% mutate(odate=floor_date(odate, "year"))

shootings_by_year <- cleaned_counts_by_race_data %>% group_by(odate, VIC_RACE) %>% summarise(y=sum(n))

ggplot(shootings_by_year, aes(x=odate, y=y, colour=VIC_RACE)) + geom_line() + stat_smooth(method="lm") + xlab("Year") + ylab("Number of shooting victims")
```

## Conclusion

In NYC, black males between the ages of 18-24 are the most likely demographic to be a shooting victim in NYC, and victims are mostly likely to be shot at night and in the early morning. This has been true for every year from 2006 to 2021. The model predicts that the number of shooting victims is on the decline. Before 2020, there was a strong steady decline in the number of shooting victims, with black victims falling the fastest. In 2020 and 2021, there was a spike among black and Hispanic victims. Further investigation using different data sources might be able to explain why these spikes occurred. 

### Possible Bias

As someone who has never experienced a shooting, and has a normal fear of being a shooting victim, my data concentrated on the victims. My bias towards the victim data was set aside as best as possible by looking at every data point about the victrims and presenting a clear representation of each.

```{r Session Info}
sessionInfo()
```